---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!--
You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

You can also embed plots in R chunks. In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gcplyr

```{r, echo = FALSE}
version <- as.vector(read.dcf('DESCRIPTION')[, 'Version'])
version <- gsub('-', '.', version)
```
<!-- badges: start -->
[![packageversion](https://img.shields.io/badge/Github-`r version`-blue.svg?style=flat&logo=github)](https://github.com/mikeblazanin/gcplyr/commits/master)
[![CRAN status](https://www.r-pkg.org/badges/version/gcplyr)](https://CRAN.R-project.org/package=gcplyr)
[![License: MIT](https://img.shields.io/badge/license-MIT-red.svg)](https://cran.r-project.org/web/licenses/MIT)
[![](http://cranlogs.r-pkg.org/badges/grand-total/gcplyr?color=yellow)](https://cran.r-project.org/package=gcplyr)
<!-- badges: end -->

## What this package can do

`gcplyr` was created to make it easier to import, wrangle, and do model-free analyses of microbial growth curve data, as commonly output by plate readers. 

* `gcplyr` can flexibly import all the common data formats output by plate readers and reshape them into 'tidy' formats for analyses.
* `gcplyr` can import experimental designs from files or directly in `R`, then merge this design information with density data.
* This merged tidy-shaped data is then easy to work with and plot using functions from `gcplyr` and popular packages `dplyr` and `ggplot2`.
* `gcplyr` can calculate plain and per-capita derivatives of density data.
* `gcplyr` has several methods to deal with noise in density or derivatives data.
* `gcplyr` can extract parameters like growth rate/doubling time, carrying capacity, diauxic shifts, extinction, and more without fitting an equation for growth to your data.

**Please send all questions, requests, comments, and bugs to [mikeblazanin@gmail.com](mailto:mikeblazanin@gmail.com)**

## Installation

You can install the most recently-released version from [GitHub](https://github.com/mikeblazanin/gcplyr/) by running the following lines in R:

``` r
install.packages("devtools")
devtools::install_github("mikeblazanin/gcplyr")
```

You can install the version most-recently released on CRAN by running the following line in R:

``` r
install.packages("gcplyr")
```

## Getting Started

The best way to get started is to check out [the online documentation](https://mikeblazanin.github.io/gcplyr/), which includes examples of all of the most common `gcplyr` functions and walks through how to import, manipulate, and analyze growth curve data using `gcplyr` from start to finish.

This documentation is also available as a series of pdf vignette files:

1. [Introduction](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gc01_gcplyr.pdf)
2. [Importing and transforming data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gc02_import_transform.pdf)
3. [Incorporating design information](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gc03_incorporate_designs.pdf)
4. [Pre-processing and plotting data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gc04_preprocess_plot.pdf)
5. [Processing data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gc05_process.pdf)
6. [Analyzing data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gc06_analyze.pdf)
7. [Dealing with noise](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gc07_noise.pdf)
8. [Statistics, merging other data, and other resources](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gc08_conclusion.pdf)
9. [Working with multiple plates](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gc09_multiple_plates.pdf)

## Citation

Please cite software as:

 Blazanin, Michael. 2023. gcplyr: an R package for microbial growth curve data analysis. bioRxiv doi: 10.1101/2023.04.30.538883.
