---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!--
You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

You can also embed plots in R chunks. In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gcplyr

<!-- badges: start -->
<!-- badges: end -->

## What this package can do

`gcplyr` facilitates import, manipulation and model-free analysis of bacterial growth curve data, as commonly output by plate readers. 

* Bacterial density time series data is commonly generated by plate readers in a number of different formats. `gcplyr` can flexibly import all the common formats and reshape them into 'tidy' formats for downstream analyses.
* Information on the experimental design of a set of growth curves might be saved in a file or may only be recorded non-digitally (e.g. in a lab notebook). `gcplyr` can import experimental designs from files or allow users to input that information directly through `gcplyr` functions, then merge this design information with previously-imported density data.
* Once data is tidy-formatted and merged with design information, it's easy to work with and plot using functions from `gcplyr` and popular packages `dplyr` and `ggplot2`.
* `gcplyr` streamlines a number of different methods to calculate derivatives of your density data.
* `gcplyr` also facilitates several strategies to deal with noise in your data, including smoothing of raw data and linear fitting during derivatives calculation.
* `gcplyr` facilitates model-free characterization and analysis of growth data to extract parameters like growth rate/doubling time, carrying capacity, diauxic shifts, extinction, and more.

**Please send all questions, requests, comments, and bugs to mikeblazanin [at] gmail [dot] com**

## Installation

You can install the latest released version from CRAN by running the following line in R:

``` r
install.packages("gcplyr")
```

If you would like to install the latest development version from [GitHub](https://github.com/mikeblazanin/gcplyr/), you can run the following lines in R:

``` r
install.packages("devtools")
devtools::install_github("mikeblazanin/gcplyr")
```

## Getting Started

The best way to get started is to check out [the online documentation](https://mikeblazanin.github.io/gcplyr/), which includes examples of all of the most common `gcplyr` functions and walks through how to import, manipulate, and analyze growth curve data using `gcplyr` from start to finish.

This documentation is also available as a series of pdf vignette files:

1. [Introduction](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gcplyr.pdf)
2. [Importing and transforming data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/import_transform.pdf)
3. [Incorporating design information](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/incorporate_designs.pdf)
4. [Pre-processing and plotting data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/preprocess_plot.pdf)
5. [Processing data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/process.pdf)
6. [Analyzing data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/analyze.pdf)
7. [Dealing with noise](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/noise.pdf)
8. [Statistics, merging other data, and other resources](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/conclusion.pdf)

## Citation

```{r, include = FALSE}
gcplyr_lib <- dirname(system.file(package = "gcplyr"))
ver <- packageDescription("gcplyr", lib.loc = gcplyr_lib)$Version
```

Please cite software as:

 Blazanin, Michael. 2023. 'gcplyr: manipulate and analyze growth curve data.' R package version `r ver`