---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!--
You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

You can also embed plots in R chunks. In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gcplyr

<!-- badges: start -->
<!-- badges: end -->

`gcplyr` facilitates import, manipulation and model-free analysis of bacterial growth curve data, as commonly output by plate readers. The package provides tools for reshaping common plate reader outputs into 'tidy' formats and merging them with design information, making data easy to work with using 'gcplyr' and other packages. Also streamlines common growth curveprocessing steps, like smoothing and calculating derivatives, and facilitates model-free characterization and analysis of growth data to extract parameters like growth rate/doubling time, carrying capacity, diauxic shifts, extinction, and more.

**Please send all questions, requests, comments, and bugs to mikeblazanin [at] gmail [dot] com**

## What this package can do

* Bacterial density time series data is commonly generated by plate readers in a number of different formats. `gcplyr` can flexibly import all the common formats and reshape them for downstream analyses.
* Information on the experimental design elements of a set of growth curves might be saved in a tabular file (e.g. an Excel file), or may only exist non-digitally (e.g. in a lab notebook). `gcplyr` can import experimental designs from files or allow users to input that information directly through `gcplyr` functions. Then, `gcplyr` can merge these design elements with the density data imported from files.
* Growth curves datasets can be processed (e.g. smoothing) and analyzed using a variety of `gcplyr` functions

## Installation

You can install gcplyr from [GitHub](https://github.com/mikeblazanin/gcplyr/) by running the following lines in R:

``` r
install.packages("devtools")
devtools::install_github("mikeblazanin/gcplyr")
```

## Getting Started

The best way to get started is to check out [the online documentation](https://mikeblazanin.github.io/gcplyr/), which includes examples of all of the most common `gcplyr` functions and walks through how to import, manipulate, and analyze growth curves data using `gcplyr` from start to finish.

This documentation is also available as a series of pdf vignette files:

1. [Introduction](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/gcplyr.pdf)
2. [Importing and transforming data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/import_transform.pdf)
3. [Incorporating design information](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/incorporate_designs.pdf)
4. [Pre-processing and plotting data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/preprocess_plot.pdf)
5. [Processing data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/process.pdf)
6. [Analyzing data](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/analyze.pdf)
7. [Dealing with noise](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/noise.pdf)
8. [Statistics, merging other data, and other resources](https://github.com/mikeblazanin/gcplyr/blob/master/vignettes/conclusion.pdf)

## Citation

```{r, include = FALSE}
gcplyr_lib <- dirname(system.file(package = "gcplyr"))
ver <- packageDescription("gcplyr", lib.loc = gcplyr_lib)$Version
```

Please cite software as:

 Blazanin, Michael. 2022. 'gcplyr: manipulate and analyze growth curve data.' R package version `r ver`