---
title: "Working with multiple plates"
author: "Mike Blazanin"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Working with multiple plates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir = tempdir())
```

# Where are we so far?

1. Introduction: `vignette("01_gcplyr")`
2. Importing and transforming data: `vignette("02_import_transform")`
3. Incorporating design information: `vignette("03_incorporate_designs")`
4. Pre-processing and plotting your data: `vignette("04_preprocess_plot")`
5. Processing your data: `vignette("05_process")`
6. Analyzing your data: `vignette("06_analyze")`
7. Dealing with noise: `vignette("07_noise")`
8. Statistics, merging other data, and other resources: `vignette("08_conclusion")`
9. **Working with multiple plates:** `vignette("09_multiple_plates")`


So far, we've covered everything you can do with `gcplyr` with a single plate. In this vignette, we'll be covering how `gcplyr` can also easily handle multiple plates of data at the same time.

If you haven't already, load the necessary packages.

```{r setup}
library(gcplyr)

library(dplyr)
library(ggplot2)
```

```{r, include = FALSE}
print_df <- function(df, col.names = FALSE) {
  write.table(format(df, justify = "right"),
              row.names=FALSE, col.names = col.names, quote = F)
}
```




