---
title: "Introduction to using gcplyr"
author: "Mike Blazanin"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Introduction to using gcplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir = tempdir())

#To re-build all pdf vignettes (only ones that are git checked-in), 
# copy-paste the code below into console (Ctrl+Enter does not work):
#for (fil in list.files(path = "./vignettes/", pattern = ".+Rmd")) {rmarkdown::render(paste0("./vignettes/", fil), rmarkdown::pdf_document(toc = TRUE, toc_depth = 4))}

#To re-build all vignettes where the rmd was changed more recently than the pdf
# copy-paste the code below into console (Ctrl+Enter does not work):
#for (fil in list.files(path = "./vignettes/", pattern = ".+Rmd")) {if(file.info(paste0("./vignettes/", fil))$mtime > file.info(paste0("./vignettes/", sub(x = fil, "\\.Rmd", ".pdf")))$mtime) {rmarkdown::render(paste0("./vignettes/", fil), rmarkdown::pdf_document(toc = TRUE, toc_depth = 4))}}

#To do so for all vignettes, copy-paste the code below into console
# (Ctrl+Enter does not work):
#for (fil in list.files(path = "./vignettes/", pattern = ".+Rmd")) {rmarkdown::render(paste0("./vignettes/", fil), rmarkdown::html_vignette(toc = TRUE, toc_dept = 4)); rmarkdown::render(paste0("./vignettes/", fil), rmarkdown::pdf_document(toc = TRUE, toc_depth = 4)); rmarkdown::render(paste0("./vignettes/", fil), rmarkdown::word_document(toc = TRUE, toc_dept = 4))}
#
#Get timings for all html builds (copy-paste not Ctrl+Enter)
#{filtimes <- data.frame(fil = list.files(path = "./vignettes/", pattern = ".+Rmd")); filtimes$times <- NA; for (filindex in 1:nrow(filtimes)) {fil <- filtimes$fil[filindex]; start <- Sys.time(); rmarkdown::render(paste0("./vignettes/", fil), rmarkdown::html_vignette(toc = TRUE, toc_dept = 4)); filtimes$times[filindex] <- difftime(Sys.time(), start, units = "mins")}; print(filtimes)}
```

# Getting started
`gcplyr` is a package that implements a number of functions to make it easier to import, manipulate, and analyze microbial growth from data collected in multiwell plate readers ("growth curves"). Without `gcplyr`, importing and analyzing plate reader data can be a complicated process that has to be tailored for each experiment, requiring many lines of code. With `gcplyr` many of those steps are now just a single line of code.

This document gives an introduction of how to use `gcplyr` for each step of a growth curve analysis.

To get started, you need your growth curve data file saved to your computer (.csv, .xls, .xlsx, or any other format that can be read by `read.table`).

Users often want to combine their data with some information on the experimental design of their plate(s). You can save this information into a tabular file as well, or you can just keep it handy to enter directly in `R` (see `vignette("gc03_incorporate_designs")`).

Let's get started by loading `gcplyr`. We're also going to load a couple other packages we'll need.

```{r setup}
library(gcplyr)

library(dplyr)
library(ggplot2)
```

# A quick demo of `gcplyr`
Before digging into the details, here's a simple demonstration of what a final `gcplyr` script can look like. This script:

1. imports data from files created by a plate reader
2. combines it with design files created by the user
3. calculates the lag time, maximum growth rate, maximum density, and area-under-the-curve

**Don't worry about understanding all the details of how the code works right now.** Each of these steps is explained in depth in later articles.

```{r}
#For the purposes of this demo, we have to create our example data and
# design files. Normally, the data would be created by a plate reader, and
# the design would be created by you, the user

#Generate our example data file, widedata.csv
make_example(vignette = 1, example = 1)

#Generate our example design files, Bacteria_strain.csv and Phage.csv
make_example(vignette = 1, example = 2)

# Read in our data 
data_wide <- read_wides(files = "widedata.csv")

# Transform our data to be tidy-shaped
data_tidy <- 
  trans_wide_to_tidy(wides = data_wide, id_cols = c("file", "Time"))

# Convert our time into hours
data_tidy$Time <- as.numeric(data_tidy$Time)/3600

# Import our designs
designs <- import_blockdesigns(files = c("Bacteria_strain.csv", "Phage.csv"))

# Merge our designs and data
data_merged <- merge_dfs(data_tidy, designs)

# Plot the data
ggplot(data = data_merged,
       aes(x = Time, y = Measurements, color = Well)) +
  geom_line(aes(lty = Phage)) + 
  guides(color = "none")

# Voila! 8 lines of code and all your data is imported & plotted!

# Calculate the per-capita growth rate over time in each well
data_merged <- mutate(
  group_by(data_merged, Well),
  percap_deriv = calc_deriv(y = Measurements, x = Time, percapita = TRUE, 
                            blank = 0, window_width_n = 5))

# Calculate four common metrics of bacterial growth:
#  the lag time, saving it to a column named lag_time
#  the maximum growth rate, saving it to a column named max_percap
#  the maximum density, saving it to a column named max_dens
#  the area-under-the-curve, saving it to a column named 'auc'
data_sum <- summarize(
  group_by(data_merged, Well, Bacteria_strain, Phage),
  lag_time = lag_time(x = Time, y = Measurements, deriv = percap_deriv),
  max_percap = max(percap_deriv, na.rm = TRUE),
  max_dens = max(Measurements),
  auc = auc(y = Measurements, x = as.numeric(Time)))

# Print some of the values
head(data_sum)

# Plot the results for each of the metrics
ggplot(data = data_sum, aes(x = lag_time)) +
  geom_histogram()

ggplot(data = data_sum, aes(x = max_percap)) +
  geom_histogram()

ggplot(data = data_sum, aes(x = max_dens)) +
  geom_histogram()

ggplot(data = data_sum, aes(x = auc)) +
  geom_histogram()
```

# What's next?

Generally, working with `gcplyr` will follow a number of steps, each of which is likely to be only one or a few lines of code in your final script. We've explained each of these steps in a page linked below. To start, we'll learn how to import our data into `R` and transform it into a convenient format.

1. Introduction: `vignette("gc01_gcplyr")`
2. **Importing and reshaping data: `vignette("gc02_import_reshape")`**
3. Incorporating experimental designs: `vignette("gc03_incorporate_designs")`
4. Pre-processing and plotting your data: `vignette("gc04_preprocess_plot")`
5. Processing your data: `vignette("gc05_process")`
6. Analyzing your data: `vignette("gc06_analyze")`
7. Dealing with noise: `vignette("gc07_noise")`
8. Best practices and other tips: `vignette("gc08_conclusion")`
9. Working with multiple plates: `vignette("gc09_multiple_plates")`
10. Using make_design to generate experimental designs: `vignette("gc10_using_make_design")`

