---
title: "Pre-processing and plotting data"
author: "Mike Blazanin"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 4
  word_document:
    toc: true
    toc_depth: 4
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Pre-processing & plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir = tempdir())

#Note: to knit to all output formats simultaneously, run in R console:
#rmarkdown::render("./vignettes/gcplyr-workflow.Rmd", output_format = "all")
```

# Pre-processing
Now that we have our data and designs merged, we're almost ready to start processing and analyzing them. However, first we need to carry out any necessary pre-processing steps, like excluding wells that were contaminated or empty, and converting time formats to numeric.

## Pre-processing: excluding data
In some cases, we want to remove some of the wells from our growth curves data before we carry on with downstream analyses. For instance, they may have been left empty, contained negative controls, or were contaminated. We can use `dplyr`'s `filter` function to remove those wells that meet criteria we want to exclude.

For instance, let's imagine that we realized that we put the wrong media into Well B1, and so we should remove it from our analyses. In that case, we can simply:

```{r}
#We have previously loaded dplyr, but if you haven't already then
#make sure to add the line:
#   library(dplyr)

example_data_and_designs_filtered <- filter(ex_dat_mrg, Well != "B1")
head(example_data_and_designs_filtered)
```

Now we can see that all rows from Well B1 have been excluded. We could do something similar if we realized that a Bacterial strain was contaminated. For instance, if strain 13 was contaminated, we could exclude it (and Well B1) as follows:

```{r}
example_data_and_designs_filtered <- 
  filter(ex_dat_mrg, 
         Well != "B1", Bacteria_strain != "Strain 13")
head(example_data_and_designs_filtered)
```


## Pre-processing: converting dates & times into numeric
Growth curve data produced by a plate reader often encodes the timestamp information as a string (e.g. "2:45:11" for 2 hours, 45 minutes, and 11 seconds), while downstream analyses need timestamp information as a numeric (e.g. number of seconds elapsed). Luckily, others have written great packages that make it easy to convert from common date-time text formats into plain numeric formats. Here, we'll see how to use `lubridate` to do so:

First we have to create a data frame with time saved as it might be by a plate reader. As usual, **don't worry how this block of code works**, since it's just creating an example file in the same format as that output by a plate reader.

```{r}
ex_dat_mrg$Time <-
  paste(ex_dat_mrg$Time %/% 3600,
        formatC((ex_dat_mrg$Time %% 3600) %/% 60, 
                width = 2, flag = 0),
        formatC((ex_dat_mrg$Time %% 3600) %% 60,
                width = 2, flag = 0),
        sep = ":")
```

Let's take a look at this data.frame. This shows the `Time` column as it might be written by a plate reader.

```{r}
head(ex_dat_mrg)
```

We can see that our `Time` aren't written in an easy numeric. Instead, they're in a format that's easy for a human to understand (but unfortunately not very usable for analysis).

Let's use `lubridate` to convert this text into a usable format. `lubridate` has a whole family of functions that can parse text with hour, minute, and/or second components. You can use `hms` if your text contains hour, minute, and second information, `hm` if it only contains hour and minute information, and `ms` if it only contains minute and second information.

Since the example has all three, we'll use `hms`. Once `hms` has parsed the text, we'll use another function to convert the output of `hms` into a pure numeric value: `time_length`. By default, `time_length` returns in units of seconds, but you can change that by changing the `unit` argument to `time_length`. See `?time_length` for details.

```{r}
#We have previously loaded lubridate, but if you haven't already then
#make sure to add the line:
#   library(lubridate)

ex_dat_mrg$Time <- time_length(hms(ex_dat_mrg$Time))

head(ex_dat_mrg)
```

And now we can see that we've gotten nice numeric `Time` values! So we can proceed with the next steps of the analysis.

# Plotting your data

Once your data has been merged and times have been converted to numeric, we can easily plot our data using the `ggplot2` package. That's because `ggplot2` was specifically built on the assumption that data would be tidy-shaped, which ours is! We won't go into depth on how to use `ggplot` here, but there are three main commands to the plot below:

* `ggplot` - the ggplot function is where you specify the `data.frame` you would like to use and the *aes*thetics of the plot (the x and y axes you would like)
* `geom_line` - tells `ggplot` how we would like to plot the data, in this case with a line (another common `geom` for time-series data is `geom_point`)
* `facet_wrap` - tells `ggplot` to plot each Well in a separate facet

**We'll be using this format to plot our data throughout the remainder of this vignette**

```{r}
#We have previously loaded ggplot2, but if you haven't already then
#make sure to add the line:
#    library(ggplot2)

#First, we'll reorder the Well levels so they plot in the correct order
ex_dat_mrg$Well <- 
  factor(ex_dat_mrg$Well,
         levels = paste(rep(LETTERS[1:8], each = 12), 1:12, sep = ""))

ggplot(data = ex_dat_mrg, aes(x = Time, y = Measurements)) +
  geom_line() +
  facet_wrap(~Well, nrow = 8, ncol = 12)
```

Generally speaking, **from here on you should plot your data frequently**, and in every way you can think of! **After every processing and analysis step, visualize both the input data and output data** to understand what the processing and analysis steps are doing and whether they are the right choices for your particular data (this vignette will be doing that too!)